Type theories as quotient inductive-recursive types

### 2025-02-14 tex/main.tex
Draft S5 and revise S4.3

\usepackage[obeyFinal,textsize=footnotesize]{todonotes}
\usepackage{manfnt}
\newcommand{\danger}{\marginpar[\hfill\dbend]{\dbend\hfill}}
\newcommand{\LT}[2][]{\todo[inline,author={L-T},caption={},color={pink},#1]{#2}}
\newcommand{\Gene}[2][]{\todo[inline,author={Gene},caption={},#1]{#2}}
\newcommand{\Fredrik}[2][]{\todo[inline,author={Fred},caption={},#1]{#2}}

@@

A typical workaround in practice \cite{Kaposi2019,Kaposi2024a} is to add equality constructors as rewrite rules after an inductive definition and further identify equality constructors with the reflxivity proof.
%
As the host theory is assumed to satisfy the uniqueness of identity proof and function extensionality, this workaround can be justified by Hofmann's result~\cite{Hofmann1996,Winterhalter2019,Kapulkin2025} that every well-typed term in extensional type theory can be translated to a term without them in intensional type theory.
%
However, this workaround is computationally ill-behaved, as confluence does not hold in general.
Also, the presentation often differs from the actual definition of type theory in intensional type theory.
%
Any method that reduces the friction introduced by the use of equality constructors without compromising on computational properties and the gap between formalisation and presentation seem worth exploring.
%

In this paper, we show that the development can be further streamlined by combining quotient inductive-inductive types and inductive-recursive types.
%
The basic idea, following a suggestion by Kaposi~\cite{Kaposi2023} and a precursor by Dybjer~\cite{Dybjer1996} and Danielsson~\cite{Danielsson2006}, is to replace the inductively defined action of a substitution on a type with a recursive definition, which computes.
%
Our development also makes use of another feature not usually found in standard presentations of type theory: recursive pattern matching definitions with overlapping patterns.
%
This is helpful to again avoid transport hell in the form of equations that do not reduce due to the standard translation of pattern matching to case trees.
%
As a result, many equations will hold strictly, and we can avoid the use of transport in the definition of type theory itself (which does not seem possible with quotient inductive-inductive types alone).
%

@@

%
We emphasise that, following a fine tradition of previous attempts to represent type theory in type theory~\cite{Danielsson2006,Chapman2009,Altenkirch2016a}, we are making use of type-theoretic features that are so far not yet well understood or foundationally well motivated.
Nevertheless, Agda can be convinced to accept our formal implementation, using some experimental flags to introduce additional rewrite rules.
Indeed, we also see this paper as a case study to motivate further developments for quotient inductive-inductive-recursive types and overlapping patterns.

\paragraph*{Contributions}
\begin{itemize}
  \item Exploration of the use of quotient inductive-inductive-recursive types and definitions by overlapping patterns.
    In particular, we give in a type theory:
    \begin{itemize}
      \item a definition of parallel substitution calculus;
      \item a definition of type theory with a universe and $\Pi$-types,
    \end{itemize}
   with their elimination principles and the standard model, 
   using quotient inductive-inductive-recursive types and definitions by overlapping patterns to define type substitution and term substitution partially.
  \item Comparison with definitions using quotient inductive-inductive types.
\end{itemize}

\subsection{Plan of the paper}

@@

We work in intensional Martin-L\"of type theory with uniqueness of identity proofs (\emph{UIP} for short). In \cref{sec:models,sec:formal-implementation}, we also assume function extensionality and $\eta$-rules for $\Pi$-types and $\Sigma$-types.
%
We mostly follow the notational conventions of the HoTT book~\cite{UFP2013} (even if we do not assume the univalence axiom), except that we write dependent function types using Agda notation $(x : A) \to B\,x$.
%
We are further inspired by Agda in writing $\implicit{x}\;B\,x$ for $(x : A) \to B\,x$ when $x$ is treated as an `implicit' argument which is not given explicitly, but instead can be figured out from context.
We write $\Set_i$ for the cumulative hierarchy of universes with the level index $i$ suppressed.

We write $x \jdeq y$ for strict equality, $x =^{A} y$ for the identity type stating that elements $x : A$ and $y : A$ of a type $A$ are equal, and $\dpath{P}{p}{t}{u}$ for the `dependent' identity type where $P : A \to \Set$, $t : P\,x$, $u : P\,y$ and $p : x =^A y$. One possible realisation of this type is as $t =^{P}_{p} u \defeq \transfib{P}{p}{t} =^{P\,y} u$, where $\transfibf{P} : (p : x =^A y) \to P\,x \to P\,y$ is our notation for the operation which allows is to transport values between type families applied to elements that are equal up to the identity type.
%
%Finally, we write $x \simeq y$ for the \emph{heterogeneous} identity type, where we allow $x : A$ and $y : B$ to have potentially different types.
%As shown by McBride~\cite{McBride1999}, the heterogeneous identity type is equivalent to the ordinary identity type equipped with the principle of uniqueness of identity proofs.

We have formally implemented our definitions and proofs in \Agda, using the options \texttt{-{}-with-K}, \texttt{-{}-exact-split}, \texttt{-{}-rewriting}, and \texttt{-{}-local-confluence-check}.
Quotient inductive types are simulated by postulating equations to introduce equality constructors~\cite{Licata2011}.
Some of our statements are meta-theoretical and thus not checked by \Agda.
Others are only partially checked, as \Agda does not generate proof obligations for postulated quotient inductive types.
We have marked such `informal' proofs with the symbol \resizebox*{!}{\baselineskip}{\textdbend}.

@@

Quotient inductive types (QITs) are a special case of higher inductive types~\cite{Lumsdaine2020} from homotopy type theory, where constructors can not only introduce new elements, but also new equalities between the introduced elements.
%
Since we assume UIP globally, all QITs are implicitly truncated to be sets in the sense of homotopy type theory.
%
For example, the type of \emph{join lists} $\JList{A}$ can be given by the QIT
with the following constructors:
\begin{alignat*}{3}
  \Jnil                    & : \implicit{A}      && \JList{A} \\
  \Jsing{\blank}           & : \implicit{A}      && A \to \JList{A} \\
  \Jconcat{\blank}{\blank} & : \implicit{A}      && \JList{A} \to \JList{A} \to \JList{A} \\
  \mathsf{idl}             & : \implicit{\ell}\; && (\Jconcat{\ell}{\Jnil}) =^{\JList{A}} \ell \\
  \mathsf{idr}             & : \implicit{\ell}\; && (\Jconcat{\Jnil}{\ell}) =^{\JList{A}} \ell \\
  \mathsf{assoc}           & : \implicit{\ell_0, \ell_1, \ell_2}\; && \Jconcat{(\Jconcat{\ell_0}{\ell_1})}{\ell_2} =^{\JList{A}} \Jconcat{\ell_0}{(\Jconcat{\ell_1}{\ell_2})}
\end{alignat*}

@@

\subparagraph*{Implementing overlapping patterns via rewrite rules}
Unfortunately, definitions by overlapping patterns are only implemented in an outdated experimental fork of \Agda.
To emulate a function by overlapping patterns in \Agda, we declare each function clause an additional rewrite rules \cite{Cockx2020,Cockx2021} (via the option \texttt{-{}-rewriting}) and use the local confluence checker (via \texttt{-{}-local-confluence-check}).
This is indeed one of the main use cases of user-defined rewrite rules discussed by Cockx~\cite{Cockx2020}. 
The caveat, though, is that the termination checker is not imposed, so we have to check the termination informally.

@@

Type substitution takes a substitution $\sigma : \Sub\;\Gamma\;\Delta$ and an inhabitant $A : \Ty\;\Delta$ to form a type under $\Gamma$, i.e., type substitution as a constructor has type
\begin{alignat*}{3}
  [\blank]\blank &: \implicit{\Gamma, \Delta} \; \Sub\;\Gamma\;\Delta \to \Ty\;\Delta \to \Ty\;\Gamma
\end{alignat*}
where $\Gamma$ and $\Delta$ are kept implicit.
\Fredrik{Could we change this to $A[\sigma]$ instead?}
\LT{I am okay with changing it in the paper: This notation is used in the formalisation to avoid using $\comp$ for substitution composition}
\LT{The rule \textsf{concat} doesn't look like concatenation with the $\tau \comp \sigma$ because the $\sigma$ is a snoc-list.}

@@

One possibility is to define the substitution calculus as a quotient inductive-inductive-recursive type, where type substitution is defined by recursion, and hence computes.
%Now that $[ \idS ]_{\Ty}\;A = A$ and $[ \sigma ; \tau ]_{\Ty}\;A = [ \sigma ]_{\Ty}\;([ \tau ]_{\Ty}\;A)$ do not make $[\blank]_{\Ty}\blank$ a (total) function, we may be tempted to define it by
%\begin{alignat*}{3}
%[\blank]_{\Ty} \blank &: \implicit{\Gamma, \Delta} \; \Sub\;\Gamma\;\Delta \to \Ty\;\Delta \to \Ty\;\Gamma \\
%[ \sigma ]_{\Ty}\;\UU & = \UU
%\end{alignat*}
Note that this does not follow the schema of inductive-recursive definitions given by Dybjer and Setzer~\cite{Dybjer2000,Dybjer2003}, since the codomain of the recursive function is the inductively defined typed itself.\footnote{%
  While the general schema for this more liberal notion of induction-recursion remains underdeveloped, this definition is accepted in \Agda and used in other formalisations of type theory~\cite{Danielsson2006}.}
%Nevertheless, $[\idS]_{\Ty}\,A$ and $[\sigma; \tau]_{\Ty}\,A$ do not reduce to~$A$ and~$[\sigma]_{\Ty}\,[\tau]_{\Ty}\,A$ respectively, even though they are provable.
%This definition fails to remove any transport in the definition of substitution calculus.
%
Since we want type substitution to both act structurally on types and commute with identities and composition, we could try to define it recursively as follows:
\begin{alignat}{3}
[\blank]_{\Ty} \blank &: \implicit{\Gamma, \Delta} \; \Sub\;\Gamma\;\Delta \to \Ty\;\Delta \to \Ty\;\Gamma \nonumber \\
[ \sigma ]_{\Ty}       \;\UU  & = \UU                                   \label{eq:type-sub-at2-1} \\
[ \idS ]_{\Ty}         \;A    & = A                                     \label{eq:type-sub-at2-2} \\
[ \sigma ; \tau ]_{\Ty}\;A    & = [ \sigma ]_{\Ty}\;([ \tau ]_{\Ty}\;A) \label{eq:type-sub-at2-3}
\end{alignat}
As explained in \cref{sec:meta:overlapping}, this definition does \emph{not} reduce \eqref{eq:type-sub-at2-2} or \eqref{eq:type-sub-at2-3} \emph{strictly}, and \eqref{eq:type-sub-at2-2} or \eqref{eq:type-sub-at2-3} will fire only if we know that $A$ is not $\UU$.
%In the above definition, only the eliminator for $\Ty\;\Delta$ will be considered using the first-match semantics, which amounts to the following definition:
%\begin{alignat*}{3}
%[\blank]_{\Ty} \blank &: \implicit{\Gamma, \Delta} \; \Sub\;\Gamma\;\Delta \to \Ty\;\Delta \to \Ty\;\Gamma \\
%[ \sigma ]_{\Ty}\;\UU  & \reduce \UU
%\end{alignat*}
%where $\reduce$ is used to emphasise the reduction (which entails the strict equality): we have done nothing different from our first attempt! 
%As a result, neither  are strict.
We may swap \eqref{eq:type-sub-at2-1} with \eqref{eq:type-sub-at2-2} and \eqref{eq:type-sub-at2-3} to make \eqref{eq:type-sub-at2-2} and \eqref{eq:type-sub-at2-3} strict, but then \eqref{eq:type-sub-at2-1} will be elaborated to 
\[
[ \emptysub ]_{\Ty}\;\UU         \judgeq \UU, \qquad
[ \sigma, t ]_{\Ty}\;\UU         \judgeq \UU, \qquad\text{and}\qquad 
[ \pi_1\,\sigma ]_{\Ty}\;\UU     \judgeq \UU,
\]
Thus in this case we have again lost the desired strict identity $[\sigma]\;\UU \equiv \UU$.
%
%making $[\sigma]\;\UU = \UU$ propositional instead.
%Not being able to reduce $[ \sigma ]_{\Ty}\;\UU$ to $\UU$ would make our later extension with a type $\El\,a$ for $a : \Tm\;\Gamma\;\UU$ require another transport in the definition.
%Likewise, for any extension with a type former such as $\Pi$-types, structural rules would still be propositional.
%That is, using the standard semantics of function definition, we are not able to make type substitution strict for \emph{both} $\Sub\;\Gamma\;\Delta$ and $\Ty\;\Delta$ at the same time.

To ensure that type substitution is strict for all %\emph
{rules}, we turn to a definition by overlapping patterns (\cref{sec:meta:overlapping}) instead to make each clause strict. 
In addition, we can include other derived rules such as $[\pi_1(\sigma, t)]_{\Ty}\;A = [ \sigma]_{\Ty}\;A$ as long as they are locally confluent and terminating.
We end up with the following definition for type substitution:
\begin{alignat}{3}
[\blank]_{\Ty} \blank            &: \implicit{\Gamma, \Delta} \; \Sub\,\Gamma\,\Delta \to \Ty\,\Delta \to \Ty\,\Gamma \nonumber \\
[ \idS ]_{\Ty}\;A                & \reduce A \label{eq:def-type-subst-1}\\
[ \sigma ; \tau ]_{\Ty}\;A       & \reduce [ \sigma ]_{\Ty}\;([ \tau ]_{\Ty}\;A) \label{eq:def-type-subst-2}\\
[ \pi_1(\sigma, t) ]_{\Ty}\;A    & \reduce [\sigma]_\Ty\;A                       \label{eq:def-type-subst-3} \\
[ \pi_1(\sigma; \tau) ]_{\Ty}\;A & \reduce [\sigma]_\Ty\;([\pi_1\tau]_\Ty\;A)    \label{eq:def-type-subst-4} \\
[ \sigma ]_{\Ty}\;\UU            & \reduce \UU                                   \label{eq:def-type-subst-5}
\end{alignat}
We first check that this definition of type substitution is confluent.\footnote{%
\Agda implements a local confluence checker with the experimental option \text{-{}-local-confluence-check}.
This option is turned on globally in our formal implementation.}

\begin{proposition}[Termination]
  Type substitution $[\blank]_{\Ty}\blank$ is terminating.
  \danger
\end{proposition}
%\begin{proof}
%  Clearly terms on the RHS are structurally smaller~\cite{Abel2002}.
%\end{proof}
\begin{proposition}[Local confluence] \label{prop:local-confluence-1}
  Type substitution $[\blank]_{\Ty}\blank$ is locally confluent.
  \danger
\end{proposition}
\begin{proof}
  To show the local confluence, we consider all local peaks, i.e.\ each combination of \eqref{eq:def-type-subst-5} with \eqref{eq:def-type-subst-1}--\eqref{eq:def-type-subst-4},
  \[
    [\idS]_{\Ty}\;\UU, \qquad [\sigma;\tau]_{\Ty}\;\UU, \qquad [\pi_1(\sigma, t)]_{\Ty}\;\UU, \qquad\text{and}\qquad [\pi_1(\sigma; \tau)]_{\Ty}\;\UU
  \]
  and show that each applicable clause does result in $\UU$.

@@

We revisit \cref{ex:pi2-comp} to demonstrate the pragmatic benefit of QIIRTs over QIITs.
\begin{example}
  The equation $\pi_2\,(\sigma; \tau) = [\sigma](\pi_2\,\tau)$ can be formally stated without any transport.
  The type of the term on the LHS is $[\pi_1(\sigma;\tau)]_{\Ty}\;A$ which is strictly equal to the type $[\sigma]_\Ty\;([\pi_1\tau]_\Ty\;A)$ of the term on the RHS.
  Moreover, the first components in each step of the proof of \eqref{eq:pi2-comp-real-proof}, are strictly equal because of \eqref{eq:def-type-subst-3} and \eqref{eq:def-type-subst-4}.
  Consequently, the back-of-the-envelope calculation~\eqref{eq:pi2-comp-proof} is \emph{formally correct}.
\end{example}

@@

Finally, we show that these simultaneously defined functions are coherent.
\begin{proposition}[Coherence]
  For every constructor $p : \sigma = \tau : \Sub\;\Gamma\;\Delta$, we have
  \danger
  \[
    \sigma \uparrow A =^{\lam{\gamma}\Sub\;(\Gamma, [ \gamma ]_{\Ty} A)\;(\Delta, A)}_{\alert{p}} \tau \uparrow A
    \qquad\text{and}\qquad
    [\sigma]_{\Tm}\,t =^{\lam{\gamma}\Tm\;\Gamma\;([\gamma]_{\Ty}A)}_{\alert{p}} [\tau]_{\Tm}\,t,
  \]
  for any type $A$ and any term $t$.
\end{proposition}
\begin{proof}
  These two identities follow from \cref{prop:correctness-2,prop:correctness-3} easily with the heterogeneous identity (see \cref{sec:meta}).
\Fredrik{This heterogeneous equality has not been used so far; it would be good to not have to introduce it all, except perhaps locally in the proof here.}
\LT{How about deferring introducing heterogeneous equality until \cref{sec:formal-implementation}?}
\end{proof}

@@

@@

  \elim_{\Sub} & : (\sigma : \Sub\;\Gamma\;\Delta) && \to \Sub^M\;(\elim_{\Ctx}\,\Gamma)\;(\elim_{\Ctx}\;\Delta)\;\sigma \\

@@

               & \elim_{\Tm}  & : (t : \Tm\;\Gamma\;A) && \to \Tm^M\;(\elim_{\Ctx}\,\Gamma)\;(\elim_{\Ty}\,A)\;t.

@@

The recursor for $(\blank)^{\succ}$ is given straightforwardly: each method $c^M_{\IR}$ is given by the corresponding constructor $c_{\I}$ in the QIIT counterpart as well as `smart constructors'.
Additional equality constructors are all derivable in the QIIT definition.

>>>>

\usepackage[obeyFinal,textsize=footnotesize]{todonotes}
\usepackage{manfnt}
\newcommand{\danger}{\marginpar[\hfill\dbend]{\dbend\hfill}}
\newcommand{\LT}[2][]{\todo[inline,author={L-T},caption={},color={pink},#1]{#2}}
\newcommand{\Gene}[2][]{\todo[inline,author={Gene},caption={},#1]{#2}}
\newcommand{\Fredrik}[2][]{\todo[inline,author={Fred},caption={},#1]{#2}}
\lstset{
    basicstyle=\large\ttfamily,
    breaklines=true,
    breakatwhitespace=true,
    literate=%
      {≡}{$\equiv$}1
      {Γ}{$\Gamma$}1
      {Δ}{$\Delta$}1 
      {→}{$\to$}1
      {≅}{$\cong$}1
}

@@

A typical workaround in practice \cite{Kaposi2019,Kaposi2024a} is to add equality constructors as rewrite rules after an inductive definition and further identify equality constructors with the reflxivity proof.
%
As the host theory is assumed to satisfy the uniqueness of identity proof and function extensionality, this workaround can be justified by conservativity of equality reflection over intensional type theory~\cite{Hofmann1996,Winterhalter2019,Kapulkin2025}.
%
However, this workaround is computationally ill-behaved, as confluence does not hold in general.
Also, the presentation often differs from the actual definition of type theory in intensional type theory.
%
Any method that reduces the friction introduced by the use of equality constructors without compromising on computational properties and the gap between formalisation and presentation seem worth exploring.
%

In this paper, we show that the development can be further streamlined by combining quotient inductive-inductive types and inductive-recursive types, called quotient inductive-inductive-recursive types (\emph{QIIRT}s for short).
%
The basic idea, following a suggestion by Kaposi~\cite{Kaposi2023} and a precursor by Dybjer~\cite{Dybjer1996} and Danielsson~\cite{Danielsson2006}, is to replace the inductively defined action of a substitution on a type with a recursive definition, which computes.
%
Our development also makes use of another feature not usually found in standard presentations of type theory: recursive pattern matching definitions with overlapping patterns.
%
This is helpful to again avoid transport hell in the form of equations that do not reduce due to the standard translation of pattern matching to case trees.
%
As a result, many equations will hold strictly, and we can avoid the use of transport in the definition of type theory itself (which does not seem possible with quotient inductive-inductive types alone).
%

@@

%
We emphasise that, following a fine tradition of previous attempts to represent type theory in type theory~\cite{Danielsson2006,Chapman2009,Altenkirch2016a}, we are making use of type-theoretic features that are so far not yet well understood or foundationally well motivated.
Nevertheless, Agda can be convinced to accept our formal implementation, using user-defined rewrite rules~\cite{Cockx2020,Cockx2021}.
Indeed, we also see this paper as a case study to motivate further developments for quotient inductive-inductive-recursive types and overlapping patterns.

\subparagraph*{Contributions}
This paper explores the use of QIIRTs and definitions by overlapping patterns to define type theories in intensional type theory.
In particular, we give
\begin{itemize}
  \item definitions of parallel substitution calculus, type theories with a universe and $\Pi$-types,
  \item with their elimination principles and the standard model;
  \item a comparison with these definitions using quotient inductive-inductive types.
\end{itemize}

\subparagraph*{Plan of the paper}
In the rest of the paper, we introduce the metatheory in \cref{sec:meta}.
Then, we will develop type theories formulated as QIIRTs in \cref{sec:QIIRTs}, followed by their elimination principles and usage in \cref{sec:models}.
We outline the \Agda formal implementation in \cref{sec:formal-implementation}, and we conclude with related and future work in \cref{sec:conclusion}.

@@

We work in intensional Martin-L\"of type theory with uniqueness of identity proofs. In \cref{sec:models,sec:formal-implementation}, we also assume function extensionality and $\eta$-rules for $\Pi$-types and $\Sigma$-types.
%
We mostly follow the notational conventions of the HoTT book~\cite{UFP2013} (even if we do not assume the univalence axiom), except that we write dependent function types using Agda notation $(x : A) \to B\,x$.
%
We are further inspired by Agda in writing $\implicit{x}\;B\,x$ for $(x : A) \to B\,x$ when $x$ is treated as an `implicit' argument which should be clear from the context.
We write $\Set_i$ for the cumulative hierarchy of universes with the level index $i$ suppressed.

We write $x \jdeq y$ for strict equality, $x =^{A} y$ for the identity type stating that elements $x : A$ and $y : A$ of a type $A$ are equal, and $\dpath{P}{p}{t}{u}$ for the `dependent' identity type where $P : A \to \Set$, $t : P\,x$, $u : P\,y$ and $p : x =^A y$. One possible realisation of this type is as $t =^{P}_{p} u \defeq \transfib{P}{p}{t} =^{P\,y} u$, where $\transfibf{P} : (p : x =^A y) \to P\,x \to P\,y$ is the operation to transport values between the type family $P$ along the identity type.
%

We have implemented our definitions and proofs in \Agda; see \cref{sec:formal-implementation} for details.
However, some statements are metatheoretical and not checked in \Agda, while others are only partially checked by \Agda, as proof obligations are given manually. %for postulated equality constructors.
We have marked such `informal' proofs with the symbol \resizebox*{!}{\baselineskip}{\textdbend}.

@@

Quotient inductive types (QITs) are a special case of higher inductive types~\cite{Lumsdaine2020} from homotopy type theory, where constructors can not only introduce new elements, but also new equalities between the introduced elements.
%
Since we assume the uniqueness of identity proofs, all QITs are implicitly truncated to be sets in the sense of homotopy type theory.
%
For example, the type of \emph{join lists} $\JList{A}$ can be given by the QIT
with the following constructors:
\begin{alignat*}{3}
  \Jnil                    & : \implicit{A}      && \JList{A} \\
  \Jsing{\blank}           & : \implicit{A}      && A \to \JList{A} \\
  \Jconcat{\blank}{\blank} & : \implicit{A}      && \JList{A} \to \JList{A} \to \JList{A} \\
  \mathsf{idl}             & : \implicit{\ell}\; && (\Jconcat{\ell}{\Jnil}) =^{\JList{A}} \ell \\
  \mathsf{idr}             & : \implicit{\ell}\; && (\Jconcat{\Jnil}{\ell}) =^{\JList{A}} \ell \\
  \mathsf{assoc}           & : \implicit{\ell_0, \ell_1, \ell_2}\; && \Jconcat{(\Jconcat{\ell_0}{\ell_1})}{\ell_2} =^{\JList{A}} \Jconcat{\ell_0}{(\Jconcat{\ell_1}{\ell_2})}
\end{alignat*}

@@

@@

Type substitution takes a substitution $\sigma : \Sub\;\Gamma\;\Delta$ and an inhabitant $A : \Ty\;\Delta$ to form a type under $\Gamma$, i.e., type substitution as a constructor has type
\begin{alignat*}{3}
  [\blank]\blank &: \implicit{\Gamma, \Delta} \; \Sub\;\Gamma\;\Delta \to \Ty\;\Delta \to \Ty\;\Gamma
\end{alignat*}
where $\Gamma$ and $\Delta$ are kept implicit.
\Fredrik[noinline]{Could we change this to $A[\sigma]$ instead?}
\LT[noinline]{I am okay with changing it in the paper: This notation is used in the formalisation to avoid using $\comp$ for substitution composition}
\LT[noinline]{The rule \textsf{concat} doesn't look like concatenation with the $\tau \comp \sigma$ because the $\sigma$ is a snoc-list.}

@@

One possibility is to define the substitution calculus as a QIIRT, where type substitution is defined by recursion, and hence computes.
%Now that $[ \idS ]_{\Ty}\;A = A$ and $[ \sigma ; \tau ]_{\Ty}\;A = [ \sigma ]_{\Ty}\;([ \tau ]_{\Ty}\;A)$ do not make $[\blank]_{\Ty}\blank$ a (total) function, we may be tempted to define it by
%\begin{alignat*}{3}
%[\blank]_{\Ty} \blank &: \implicit{\Gamma, \Delta} \; \Sub\;\Gamma\;\Delta \to \Ty\;\Delta \to \Ty\;\Gamma \\
%[ \sigma ]_{\Ty}\;\UU & = \UU
%\end{alignat*}
Note that this does not follow the schema of inductive-recursive definitions given by Dybjer and Setzer~\cite{Dybjer2000,Dybjer2003}, since the codomain of the recursive function is the inductively defined typed itself.\footnote{%
  While the general schema for this more liberal notion of induction-recursion remains underdeveloped, this definition is accepted in \Agda and used in other formalisations of type theory~\cite{Danielsson2006}.}
%Nevertheless, $[\idS]_{\Ty}\,A$ and $[\sigma; \tau]_{\Ty}\,A$ do not reduce to~$A$ and~$[\sigma]_{\Ty}\,[\tau]_{\Ty}\,A$ respectively, even though they are provable.
%This definition fails to remove any transport in the definition of substitution calculus.
%
Since we want type substitution to both act structurally on types and commute with identities and composition, we could try to define it recursively as follows:
\begin{alignat}{3}
[\blank]_{\Ty} \blank &: \implicit{\Gamma, \Delta} \; \Sub\;\Gamma\;\Delta \to \Ty\;\Delta \to \Ty\;\Gamma \nonumber \\
[ \sigma ]_{\Ty}       \;\UU  & = \UU                                   \label{eq:type-sub-at2-1} \\
[ \idS ]_{\Ty}         \;A    & = A                                     \label{eq:type-sub-at2-2} \\
[ \sigma ; \tau ]_{\Ty}\;A    & = [ \sigma ]_{\Ty}\;([ \tau ]_{\Ty}\;A) \label{eq:type-sub-at2-3}
\end{alignat}
As explained in \cref{sec:meta:overlapping}, this definition does \emph{not} reduce \eqref{eq:type-sub-at2-2} or \eqref{eq:type-sub-at2-3} \emph{strictly}, and \eqref{eq:type-sub-at2-2} or \eqref{eq:type-sub-at2-3} will fire only if we know that $A$ is not $\UU$.
%In the above definition, only the eliminator for $\Ty\;\Delta$ will be considered using the first-match semantics, which amounts to the following definition:
%\begin{alignat*}{3}
%[\blank]_{\Ty} \blank &: \implicit{\Gamma, \Delta} \; \Sub\;\Gamma\;\Delta \to \Ty\;\Delta \to \Ty\;\Gamma \\
%[ \sigma ]_{\Ty}\;\UU  & \reduce \UU
%\end{alignat*}
%where $\reduce$ is used to emphasise the reduction (which entails the strict equality): we have done nothing different from our first attempt! 
%As a result, neither  are strict.
We may swap \eqref{eq:type-sub-at2-1} with \eqref{eq:type-sub-at2-2} and \eqref{eq:type-sub-at2-3} to make \eqref{eq:type-sub-at2-2} and \eqref{eq:type-sub-at2-3} strict, but then \eqref{eq:type-sub-at2-1} will be elaborated to 
\[
[ \emptysub ]_{\Ty}\;\UU         \judgeq \UU, \qquad
[ \sigma, t ]_{\Ty}\;\UU         \judgeq \UU, \qquad\text{and}\qquad 
[ \pi_1\,\sigma ]_{\Ty}\;\UU     \judgeq \UU,
\]
Thus in this case we have again lost the desired strict identity $[\sigma]\;\UU \equiv \UU$.
%
%making $[\sigma]\;\UU = \UU$ propositional instead.
%Not being able to reduce $[ \sigma ]_{\Ty}\;\UU$ to $\UU$ would make our later extension with a type $\El\,a$ for $a : \Tm\;\Gamma\;\UU$ require another transport in the definition.
%Likewise, for any extension with a type former such as $\Pi$-types, structural rules would still be propositional.
%That is, using the standard semantics of function definition, we are not able to make type substitution strict for \emph{both} $\Sub\;\Gamma\;\Delta$ and $\Ty\;\Delta$ at the same time.

To ensure that type substitution is strict for all %\emph
{rules}, we turn to a definition by overlapping patterns (\cref{sec:meta:overlapping}) instead to make each clause strict. 
In addition, we can include other derived rules such as $[\pi_1(\sigma, t)]_{\Ty}\;A = [ \sigma]_{\Ty}\;A$ as long as they are locally confluent and terminating.
We end up with the following definition for type substitution:
\begin{alignat}{3}
[\blank]_{\Ty} \blank            &: \implicit{\Gamma, \Delta} \; \Sub\,\Gamma\,\Delta \to \Ty\,\Delta \to \Ty\,\Gamma \nonumber \\
[ \idS ]_{\Ty}\;A                & \reduce A \label{eq:def-type-subst-1}\\
[ \sigma ; \tau ]_{\Ty}\;A       & \reduce [ \sigma ]_{\Ty}\;([ \tau ]_{\Ty}\;A) \label{eq:def-type-subst-2}\\
[ \pi_1(\sigma, t) ]_{\Ty}\;A    & \reduce [\sigma]_\Ty\;A                       \label{eq:def-type-subst-3} \\
[ \pi_1(\sigma; \tau) ]_{\Ty}\;A & \reduce [\sigma]_\Ty\;([\pi_1\tau]_\Ty\;A)    \label{eq:def-type-subst-4} \\
[ \sigma ]_{\Ty}\;\UU            & \reduce \UU                                   \label{eq:def-type-subst-5}
\end{alignat}
We first check that this definition of type substitution is confluent.
\begin{proposition}[Termination]
  Type substitution $[\blank]_{\Ty}\blank$ is terminating.
  \danger
\end{proposition}
%\begin{proof}
%  Clearly terms on the RHS are structurally smaller~\cite{Abel2002}.
%\end{proof}
\begin{proposition}[Local confluence] \label{prop:local-confluence-1}
  Type substitution $[\blank]_{\Ty}\blank$ is locally confluent.
  \danger
\end{proposition}
\begin{proof}
  To show the local confluence, we consider all local peaks, i.e.\ each combination of \eqref{eq:def-type-subst-5} with \eqref{eq:def-type-subst-1}--\eqref{eq:def-type-subst-4},
  \[
    [\idS]_{\Ty}\;\UU, \qquad [\sigma;\tau]_{\Ty}\;\UU, \qquad [\pi_1(\sigma, t)]_{\Ty}\;\UU, \qquad\text{and}\qquad [\pi_1(\sigma; \tau)]_{\Ty}\;\UU
  \]
  and show that each applicable clause does result in $\UU$.

@@

We revisit \cref{ex:pi2-comp} to demonstrate the pragmatic benefit of QIIRTs over QIITs.
\begin{example} \label{ex:pi-comp-qiirt}
  The equation $\pi_2\,(\sigma; \tau) = [\sigma](\pi_2\,\tau)$ can be formally stated without any transport.
  The type of the term on the LHS is $[\pi_1(\sigma;\tau)]_{\Ty}\;A$ which is strictly equal to the type $[\sigma]_\Ty\;([\pi_1\tau]_\Ty\;A)$ of the term on the RHS.
  Moreover, the first components in each step of the proof of \eqref{eq:pi2-comp-real-proof}, are strictly equal because of \eqref{eq:def-type-subst-3} and \eqref{eq:def-type-subst-4}.
  Consequently, the back-of-the-envelope calculation~\eqref{eq:pi2-comp-proof} is \emph{formally correct}.
\end{example}

@@

Finally, we show that these simultaneously defined functions are coherent.
\begin{proposition}[Coherence]
  For every constructor $p : \sigma = \tau : \Sub\;\Gamma\;\Delta$, we have
  \danger
  \[
    \sigma \uparrow A =^{\lam{\gamma}\Sub\;(\Gamma, [ \gamma ]_{\Ty} A)\;(\Delta, A)}_{\alert{p}} \tau \uparrow A
    \qquad\text{and}\qquad
    [\sigma]_{\Tm}\,t =^{\lam{\gamma}\Tm\;\Gamma\;([\gamma]_{\Ty}A)}_{\alert{p}} [\tau]_{\Tm}\,t,
  \]
  for any type $A$ and any term $t$.
\end{proposition}
\begin{proof}
  These two identities follow from \cref{prop:correctness-2,prop:correctness-3} easily with the heterogeneous identity (see \cref{sec:meta}).
\end{proof}

@@

\begin{alignat*}{10}

@@

  \elim_{\Sub} & : (\sigma : \Sub\;\Gamma\;\Delta) && \to \Sub^M\; && (\elim_{\Ctx}\,\Gamma)\;(\elim_{\Ctx}\;\Delta)\;&&\sigma \\

@@

               & \elim_{\Tm}  & : (t : \Tm\;\Gamma\;A) && \to \Tm^M\; && (\elim_{\Ctx}\,\Gamma)\;(\elim_{\Ty}\,A)\;&&t.

@@

The recursor for $(\blank)^{\succ}$ is given straightforwardly: each method $c^M_{\IR}$ is given by the corresponding constructor $c_{\I}$ in the QIIT counterpart.
Additional equality constructors are all derivable in the QIIT definition.

On the other hand, the converse $(\blank)^{\prec}$ has some choices.
Since in the QIIT definition there is only one term substitution $[\blank]_{\I}\blank$ and lifting $(\blank)^+_{\I}$, we may choose to translate $[\sigma]_{\I} \,t$ to the recursive form $[\sigma^{\prec}]_{\Tm}\,t^{\prec}$ to carry out some computation, or to the constructor form $[\sigma^{\prec}]_{\IR}\,t^{\prec}$ to stay verbatim.
Either choice is possible, but we need to use \cref{prop:correctness-1,prop:correctness-3} to switch from one to the other whenever suitable.
For example, the equality constructor $\mathsf{concat}_{\IR}$ uses the constructor form, so the method $\mathsf{concat}^M_{\I}$ relies on the identity $[ \sigma ]\,t = [\sigma]_{\Tm}\,t$ if the recursive form $[\sigma^{\prec}]_{\Tm}\,t^{\prec}$ is chosen for the translation.
